---
title: "DS Capstone Yelp Dataset - Relationship between Positive, Negative, and Profane
  Words in Reviews of Restaurants and Rating (Stars) Given"
author: "Andre Johnson"
date: "November 14, 2015"
output: word_document
---

#**Introduction and Overview**#
  
Yelp was founded in 2004 to help people find great local businesses like restaurants, dentists, hair stylists and mechanics. Yelp had a monthly average of 83 million unique visitors who visited Yelp via their mobile device in Q2 2015*.  Yelp has provided a subset of data from its users as part of a Dataset Challenge.   This set includes information about local businesses in 10 cities across 4 countries.

The intention of this project is to utilize the reviews data provided in the Yelp Dataset for restaurants and explore the relationship between the quantity/ratios of positive versus negative and profane words used in reviews of restaurants in the dataset and the rating (number of stars) given.  Furthermore we will attempt to develop the start of a model that evaluates the text of a review to predict the rating to be given.  

##**Setting up R Enviornment**##

Prepare the R enviornment for analysis.

```{r}
set.seed(1973)

require(ggplot2)
library(plyr)
require(tm)
require("knitr")
opts_chunk$set(cache=TRUE, cache.path = 'test_files/', fig.path='figure/')
library(data.table)
library(stringr)
require(stringi)
library(caret)
require(rpart)
require(rpart.plot)
library(randomForest)
library(corrplot)
library(psych)
require(descr)
```

##**Obtaining the data**## "- Download the data and load/manipulate it in R

**Task 1 - Data acquisition and cleaning**

"The goal of this task is to get familiar with the databases and do the necessary cleaning. After this exercise, we should understand what real data looks like and how much effort into cleaning the data.
"
**Download File**
  
  ```{r}
#setwd("E:/User/Coursera/Capstone/Yelp Dataset Challenge/Data")
#for simplicity, already saved data as rds and read rds file as required
#url <- "https://d396qusza40orc.cloudfront.net/dsscapstone/dataset/yelp_dataset_challenge_academic_dataset.zip"
#if (!file.exists("yelp_dataset_challenge_academic_dataset.zip")){
 # download.file(url, "yelp_dataset_challenge_academic_dataset.zip")
  #unzip("yelp_dataset_challenge_academic_dataset.zip")
#}
```

**Read Files**

  
```{r}
#yelpReviewData <- fromJSON(paste("[",paste(readLines("yelp_academic_dataset_review.json"), collapse=","),"]"))
#yelpBusinessData <- fromJSON(paste("[",paste(readLines("yelp_academic_dataset_business.json"), collapse=","),"]"))
#convert the data from JSON to R readable format
#saveRDS(yelpReviewData, "yelpReviewData.rds")
#saveRDS(yelpBusinessData, "yelpBusinessData.rds")
#save data to rds for future use

#for simplicity, saved data as rds and read rds file as required

yelpReviewData <- readRDS("yelpReviewData.rds")
yelpBusinessData <- readRDS("yelpBusinessData.rds")

```

**Create Sample File** - Given the large size of the file, we will create a sample file of randomly selected lines from the yelpReviewData.

```{r}
yelpReviewData <- yelpReviewData[,-c(1:3, 5, 7)]  
#Delete unnecessary columns in yelpReviewData - everything but text, stars, and business_id columns

yelpReviewSample <- yelpReviewData[sample(nrow(yelpReviewData), size = 2000, replace = FALSE, prob = NULL),]
#yelpReviewSample <- yelpReviewData[sample(nrow(yelpReviewData), size = 25000, replace = FALSE, prob = NULL),]
#create a sample of reviews data to work with
rm(yelpReviewData)
# remove yelpReviewData to save working memory

```

**Merge with Data from subset of yelpBusinessData for Restaurants**

```{r}
yelpRestaurantBusinessData <- subset(yelpBusinessData, grepl("Restaurants", yelpBusinessData$categories))
#subset yelpBusinessData into only restaurants
rm(yelpBusinessData) #remove yelpBusinessData to save working memory
yelpRestaurantBusinessData <- yelpRestaurantBusinessData[, -c(2:4, 6:15)]
#delete columns not necessary for analysis
yelpRestaurantBusinessData$categories <- "Restaurants"
#replace categories with simply "Restaurants"

yelpReviewRestData <- merge(yelpReviewSample, yelpRestaurantBusinessData, sort=FALSE)
#merge two datasets
rm(yelpReviewSample)
rm(yelpRestaurantBusinessData)
# remove unneeded datasets to save working memory
yelpReviewRestData <- yelpReviewRestData[, -c(1, 4)]
#remove business_id and categories columns as no longer necessary

```

**Clean up yelpReviewRestData**

```{r} 
yelpReviewRestData$text <- tolower(yelpReviewRestData$text)
#change all letters to lower case
yelpReviewRestData$text <- str_replace_all(yelpReviewRestData$text, "not ", "not" ) 
#remove space after not to convert to a positive or negative word

yelpReviewRestData$text <- str_replace_all(yelpReviewRestData$text, "don't ", "don't" ) 
#remove space after don't to convert to a positive or negative word

yelpReviewRestData$text <- str_replace_all(yelpReviewRestData$text, "nothing ", "nothing" ) 
#remove space after nothing to convert to a positive or negative word

yelpReviewRestData$text <- str_replace_all(yelpReviewRestData$text, "never ", "never" ) 
#remove space after never to convert to a positive or negative word

trimws(yelpReviewRestData$text)
#clean up text to get rid of extra whitespace

yelpReviewRestData$text <- gsub("[[:punct:]]", "", yelpReviewRestData$text)
#remove punctuation

```

**Load positive, negative, and profane word lists**

"The lists of positive, negative, and profane lists have been updated throughout the process to try and capture positive, negative, 
and profane aspects of the reviews as much as possible which over time has improved the model significantly."

```{r}
positivewords <- readLines("positivewords.txt", skipNul = TRUE)
negativewords <- readLines("negativewords.txt", skipNul = TRUE)
profanity <- readLines("profanity.txt", skipNul = TRUE)
```

**Populate columns with number of positive, negative, profane, total words and percentages**

```{r}
yelpReviewRestData$`Count Positive words` <- lapply(yelpReviewRestData$text, function(x) sum(stri_detect_fixed(x, positivewords)))
#count of positive words

yelpReviewRestData$`Count Negative Words` <- lapply(yelpReviewRestData$text, function(x) sum(stri_detect_fixed(x, negativewords)))
#count of negative words
yelpReviewRestData$`Count Profane Words` <- lapply(yelpReviewRestData$text, function(x) sum(stri_detect_fixed(x, profanity)))
#count of profane words
yelpReviewRestData$`Total Words` <- lapply(yelpReviewRestData$text, function(x) length(strsplit(x, ' ')[[1]]))
#count of total words

yelpReviewRestData$`Count Positive words` <- as.numeric(yelpReviewRestData$`Count Positive words`)
yelpReviewRestData$`Count Negative Words` <- as.numeric(yelpReviewRestData$`Count Negative Words`)
yelpReviewRestData$`Count Profane Words` <- as.numeric(yelpReviewRestData$`Count Profane Words`)
yelpReviewRestData$`Total Words` <- as.numeric(yelpReviewRestData$`Total Words`)
yelpReviewRestData$stars <- as.numeric(yelpReviewRestData$stars)
#convert entries to numeric

rm(negativewords)
rm(positivewords)
rm(profanity)

yelpReviewRestData$`Percentage of Pos Words to Total` <- round((yelpReviewRestData$`Count Positive words`/ yelpReviewRestData$`Total Words`) * 100)
yelpReviewRestData$`Percentage of Neg Words to Total` <- round((yelpReviewRestData$`Count Negative Words`/ yelpReviewRestData$`Total Words`) * 100)
#create columns for round of percentage of positive and negative words to total words in review

yelpReviewRestData$`Count Negative Words` <- str_replace_all(yelpReviewRestData$`Count Negative Words`, "0", "1" )
#change negative words count where 0 to 1 so that will not result in Inf
yelpReviewRestData$`Count Negative Words` <- as.numeric(yelpReviewRestData$`Count Negative Words`)

yelpReviewRestData$`Ratio of Pos to Neg Words` <- yelpReviewRestData$`Count Positive words`/ yelpReviewRestData$`Count Negative Words`
#create column for ratio of positive to negative words

yelpReviewRestData$`Ratio of Pos to Neg Words`[yelpReviewRestData$`Ratio of Pos to Neg Words` < 1] <- 0
#Change Ratio of Pos to Neg Words where less than 1 to 0 to bucket Negative words exceeding positive words as one factor


yelpReviewRestData <- yelpReviewRestData[, -c(2:4, 6)]
#delete text, count positive words, count negative words, and total words columns as no longer necessary and redundant with other factors

yelpReviewRestDataTrain <- yelpReviewRestData

yelpReviewRestDataTrain$`Ratio of Pos to Neg Words`[yelpReviewRestDataTrain$`Ratio of Pos to Neg Words` > 3] <- "Very Positive"
yelpReviewRestDataTrain$`Ratio of Pos to Neg Words`[yelpReviewRestDataTrain$`Ratio of Pos to Neg Words` <= 3 & yelpReviewRestData$`Ratio of Pos to Neg Words` >=1] <- "Positive"
yelpReviewRestDataTrain$`Ratio of Pos to Neg Words`[yelpReviewRestDataTrain$`Ratio of Pos to Neg Words` < 1] <- "Negative"
#Bucket Ratios into Negative, Positive, and Very Positive

yelpReviewRestDataTrain$`Percentage of Pos Words to Total`[yelpReviewRestDataTrain$`Percentage of Pos Words to Total` > 5] <- "Very Positive"
yelpReviewRestDataTrain$`Percentage of Pos Words to Total`[yelpReviewRestDataTrain$`Percentage of Pos Words to Total` <= 5 & yelpReviewRestData$`Percentage of Pos Words to Total` >=2] <- "Positive"
yelpReviewRestDataTrain$`Percentage of Pos Words to Total`[yelpReviewRestDataTrain$`Percentage of Pos Words to Total` < 2] <- "Negative"
#Bucket Ratios into Negative, Positive, and Very Positive

yelpReviewRestDataTrain$`Percentage of Neg Words to Total`[yelpReviewRestDataTrain$`Percentage of Neg Words to Total` > 3] <- "Very Negative"
yelpReviewRestDataTrain$`Percentage of Neg Words to Total`[yelpReviewRestDataTrain$`Percentage of Neg Words to Total` <= 3 & yelpReviewRestData$`Percentage of Neg Words to Total` >=1] <- "Negative"
yelpReviewRestDataTrain$`Percentage of Neg Words to Total`[yelpReviewRestDataTrain$`Percentage of Neg Words to Total` < 1] <- "Positive"
#Bucket Ratios into Negative, Positive, and Very Positive

```

"After testing the model against individual stars, it was determined that predicability of the model was not nuanced enough 
to determine whether a 4 or 5 stars, 3 stars, or 1 or 2 star so we decided to bucket the 4 and 5 stars as Good, 1 and 2 stars as Bad and 3 stars as Average
to try and improve the predictability in smaller buckets with the possibility of expanding back to 1-5 stars once further enhancement to review of text and 
the model has been done."

```{r}
yelpReviewRestDataTrain$stars <- str_replace_all(yelpReviewRestDataTrain$stars, "5", "Good" ) 
#bucket 4 and 5 stars as "Good"

yelpReviewRestDataTrain$stars <- str_replace_all(yelpReviewRestDataTrain$stars, "4", "Good" ) 
#bucket 4 and 5 stars as "Good"

yelpReviewRestDataTrain$stars <- str_replace_all(yelpReviewRestDataTrain$stars, "3", "Average" ) 
#bucket 3 stars as "Average"

yelpReviewRestDataTrain$stars <- str_replace_all(yelpReviewRestDataTrain$stars, "2", "Bad" ) 
#bucket 2 and 1 stars as "Bad"

yelpReviewRestDataTrain$stars <- str_replace_all(yelpReviewRestDataTrain$stars, "1", "Bad" ) 
#bucket 2 and 1 stars as "Bad"

yelpReviewRestDataTrain$stars <- as.factor(yelpReviewRestDataTrain$stars)
yelpReviewRestDataTrain$`Count Profane Words` <- as.factor(yelpReviewRestDataTrain$`Count Profane Words`)
yelpReviewRestDataTrain$`Ratio of Pos to Neg Words` <- as.factor(yelpReviewRestDataTrain$`Ratio of Pos to Neg Words`)
yelpReviewRestDataTrain$`Percentage of Pos Words to Total` <- as.factor(yelpReviewRestDataTrain$`Percentage of Pos Words to Total`)
yelpReviewRestDataTrain$`Percentage of Neg Words to Total` <- as.factor(yelpReviewRestDataTrain$`Percentage of Neg Words to Total`)
#convert to factors for use in model
```

##**Task 2 - Exploratory analysis**##

"The first step in building a predictive model for text is understanding the distribution and relationship between the types of words used and the rating.  
The goal of this task is to understand the basic relationships observed in the data and prepare to build a predictive model.

We will perform a thorough exploratory analysis of the data, understanding the distribution of positive, negative and profane words and relationship between the words and the rating.

First, we look at frequency of stars, positive to negative, positive and negative words to total, and total profane words in the data.
"
```{r}
summary(yelpReviewRestDataTrain)
table(yelpReviewRestData$stars, yelpReviewRestDataTrain$`Count Profane Words`)
table(yelpReviewRestData$stars, yelpReviewRestDataTrain$`Percentage of Pos Words to Total`)
table(yelpReviewRestData$stars, yelpReviewRestDataTrain$`Percentage of Neg Words to Total`)
table(yelpReviewRestData$stars, yelpReviewRestDataTrain$`Ratio of Pos to Neg Words`)

```
**Figure 4** in the Appendix will show a histogram and cross tabulation plots of the different factors and 
**Figure 5** shows box plots of Stars vs the other factors.

The cross tabulation tables and box plots show a clear trend of higher positive to negative ratio, less negative words, and less profanity in text where the stars are 4 or 5. And the opposite trend of lower positive to negative ratio, more negative words, and more profanity in text where the stars are 1 or 2 which gives hope that a model can be developed to provide some predicability in the stars to be given by analysis of the text written.


##**Methods - Task 3 - Develop a Model**##

Having cleaned up the data which now includes 5 factors we are going to develop a model using *Random Forest* technique to allow us to predict the number of stars to be given based on the text submitted. 

"randomForest implements Breiman's random forest algorithm (based on Breiman and Cutler's original Fortran code) for classification and regression. It can also be used in unsupervised mode for assessing proximities among data points."


#**Create new Training and Testing data sets**#

First, with the data cleaned up in the "yelpReviewRestDataTrain" set, we create a training and test set.

```{r}
inTrain = createDataPartition(y=yelpReviewRestDataTrain$star, p=0.7, list=FALSE)
train = yelpReviewRestDataTrain[inTrain,]
test = yelpReviewRestDataTrain[-inTrain,]
dim(train)
#rm(yelpReviewRestDataTrain)
```

#**Developing a Model**#

Now that we have a train and test data set we are going to develop a model using *Random Forest* technique to allow us to predict the number of stars 
to be given based on the text submitted. We will use a 10-fold cross validation in the model.    

```{r}

RFcontrol <- trainControl(method="cv", number=10)

RFmodel = train(stars ~ ., method="rf", data=train, trControl=RFcontrol, ntree = 250)
saveRDS(RFmodel, "rfmodel.RDS")

#RFmodel <- readRDS("rfmodel.RDS")
RFmodel

```

Maximum accuracy is at mtry = 7 and is 66.56% with out of sample error of 33.44% which is a relatively good start.  

Below is a plot of the random forest model.

**Figure 1 - Plot of RFmodel**
  
```{r}
plot(RFmodel)
```

#**Testing and Evaluating the Model**#

Now that we have created a model against the training data, lets test it against the test data set.

```{r}
confusionMatrix(test$stars, predict(RFmodel, test))
RFModelAccuracy <- mean(predict(RFmodel, test) == test$stars) * 100
```

The estimated accuracy of the model is **`r RFModelAccuracy`%** when applied to the test data set with an out of sample error of **`r 100 - RFModelAccuracy`%**.  A plot of each variable is below.

**Figure 2 - Plot of variable importance of RFModel**

```{r}

plot(varImp(RFmodel))

```
**Figure 3 - Plot of Decision Tree model**
  
```{r}

decTree <- rpart(train$stars ~ ., data=train, method="class")
prp(decTree)
```

#**Results - Conclusion**#

A random forest trainer appears to be a great start in developing a model to predict against the data sets with an out of sample of **`r 100 - RFModelAccuracy`%**.  
Further work will be required to refine the model and the predicability, in particular a review of the positivewords, negativewords, and profanity sets to be able 
to more accurately capture positive and negative sentiments and nuances in the text and data.  

As is seemingly apparent in *Figures 4 and 5* in the Appendix, there is a relationship between the number of stars given and the quantity/ratio of 
positive vs negative words and profane words in the text of the review.  While the model ultimately requires further refinement to improve predicability, 
clearly there is a relationship that exists in the text and the stars given.  

##**Discussion - Summary and Next Steps**##

This capstone project is the first step in understanding/preprocessing/exploring the data and developing a model that will be able to predict the rating with some consistency
based on the text in the review.

The work done so far is a good step in the right direction having achieved a 66% predicability level but further work will be required to refine the model to improve predicability, 
in particular a review of the positivewords, negativewords, and profanity sets to be able to more accurately capture positive and negative sentiments and nuances in the text and data.


#**Appendices**#

**Figure 4 - Plots of Stars and Cross Tabulation Tables of Stars vs Other Factors**

```{r}
par(mfrow=c(3,2))
hist(yelpReviewRestData$stars, xlab="stars", main = "Histogram of Frequency of Stars", col = (c("lightblue","gray")))
crosstab(yelpReviewRestData$stars, yelpReviewRestDataTrain$`Count Profane Words`, xlab="Count of Profane Words", ylab="Stars", main="Cross Table", col = (c("lightblue","gray")))
crosstab(yelpReviewRestData$stars, yelpReviewRestDataTrain$`Percentage of Pos Words to Total`, xlab="Positive Words to Total", ylab="Stars", main="Cross Table", col = (c("lightblue","gray")))
crosstab(yelpReviewRestData$stars, yelpReviewRestDataTrain$`Percentage of Neg Words to Total`, xlab="Negative Words to Total", ylab="Stars", main="Cross Table", col = (c("lightblue","gray")))
crosstab(yelpReviewRestData$stars, yelpReviewRestDataTrain$`Ratio of Pos to Neg Words`, xlab="Ratio of Positive to Negative Words", ylab="Stars", main="Cross Table", col = (c("lightblue","gray")))

```

**Figure 5 - Boxplots of Stars vs Other Factors**
  
```{r}
par(mfrow=c(2,2))
boxplot(yelpReviewRestData$star ~ yelpReviewRestData$`Ratio of Pos to Neg Words`, col = (c("lightblue","gray")), ylab = "Stars", xlab = "Ratio of Positive to Negative Words", main = "Box plot Stars vs Ratio of Positive to Negative Words" )
boxplot(yelpReviewRestData$star ~ yelpReviewRestData$`Percentage of Pos Words to Total`, col = (c("lightblue","gray")), ylab = "Stars", xlab = "Percentage of Pos Words to Total", main = "Box plot Stars vs Pos Words to Total" )
boxplot(yelpReviewRestData$star ~ yelpReviewRestData$`Percentage of Neg Words to Total`, col = (c("lightblue","gray")), ylab = "Stars", xlab = "Percentage of Neg Words to Total", main = "Box plot Stars vs Neg Words to Total" )
boxplot(yelpReviewRestData$star ~ yelpReviewRestData$`Count Profane Words`, col = (c("lightblue","gray")), ylab = "Stars", xlab = "Count Profane Words", main = "Box plot Stars vs Count Profane Words" )
```



**References:**
**http://www.yelp.com/about
**http://www.enchantedlearning.com/wordlist/negativewords.shtml
**http://www.enchantedlearning.com/wordlist/positivewords.shtml
**Banned Words List:** http://www.bannedwordlist.com/lists/swearWords.txt
https://en.wikipedia.org/wiki/N-gram
**http://www.inside-r.org/packages/cran/randomforest/docs/randomforest
**Natural language processing Wikipedia page:** http://en.wikipedia.org/wiki/Natural_language_processing
**Text mining infrastucture in R:** http://www.jstatsoft.org/v25/i05/
**CRAN Task View: Natural Language Processing:** http://cran.r-project.org/web/views/NaturalLanguageProcessing.html
**Basic Text Mining in R:** https://rstudio-pubs-static.s3.amazonaws.com/31867_8236987cf0a8444e962ccd2aec46d9c3.html